Experience Report: Haskell in Computational Biology

### 2012-07-01 mrfy_experience_report.tex
new graphs
more work on graphs
possible bug in exposition

\begin{document}
\parskip=0.9\parskip % cheating
\advance \parskip by 0pt plus 0.1pt

@@

When a real protein folds in three dimensions, 
amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
Some proteins contain groups of such acids called \emph{beta
strands}.
Beta strands
can become hydrogen-bonded to each other,
making them ``stuck together,''
and these beta strands help distinguish groups of homologous
proteins from one another.
MRFy detects homologous proteins that include hydrogen-bonded beta
strands; using prior methods, many instances of this problem are
intractable.

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.

@@

\begin{figure}[h!]
\centerline{\includegraphics[width=6cm]{mrfy_multicore_log}}
\caption{MRFy run-time performance with varying
numbers of processing cores on a large protein}\label{multicore}
\end{figure}

>>>>

\begin{document}
%\parskip=0.9\parskip % cheating
%\advance \parskip by 0pt plus 0.1pt

@@

We're using our implementation of Viterbi's algorithm to help detect
homologies in new kinds of proteins.
When a real protein folds in three dimensions, 
amino acids 
that are far away in the one-dimensional sequence can be
adjacent in three-dimensional space.
Some proteins contain groups of such acids called \emph{beta
strands}.
Beta strands
can become hydrogen-bonded to each other,
making them ``stuck together,''
and these beta strands help distinguish groups of homologous
proteins from one another.
MRFy detects homologous proteins that include hydrogen-bonded beta
strands; using prior methods, many instances of this problem are
intractable.

To~account for beta strands, computational biologists add
a term to Viterbi's equations, and they build
richer models of protein structure.
%%  The~richer model recognizes that only certain pairs of amino acids are
%%  able to bond to each other.
%%  The presence of beta strands changes both the natu
%%   the probability that a given
%%  query sequence fit
%%  A~mutation in a paired beta strand may break the pairing, in which
%%   case the mutated protein will not fold properly and is unlikely to
%%   survive natural selection.
%%  But a mutation in a paired beta strand \emph{can} survive if it is
%%   accompanied by a \emph{compensatory} mutation in the strand with which it is
%%   paired. 
When column~$j$ of an alignment is part of a beta strand and is paired
with another column  $\pairedwith j$,
the probability of finding amino acid~$x_i$ in column~$j$ 
depends on~$x_{\pairedwith j}$.
\remark{This looks wrong.  $x$ should be subscripted with a
query-sequence index, but $\pairedwith j$ is a model index.}
The~relevant term in
Viterbi's equations 
$V_{j}^{\prime M}(i)$ becomes
$$W_{j}^{M}(i) = V_{j}^{\prime M}(i) - \log P(x_{i} \mid x_{\pairedwith j}).$$
Columns~$j$ and $\pairedwith j$ can be as close together as a few
columns or as far apart as a few hundreds of columns.
Because $W_j^M(i)$~depends not only on $x_i$ and its local neighbors but
also on the nonlocal $x_{\pairedwith j}$, 
dynamic programming cannot compute the maximum likelihood quickly
\cite{Menke:2010ti,Daniels:2012}.

@@

%%  \begin{figure}[h!]
%%  \centerline{\includegraphics[width=6cm]{mrfy_multicore_log}}
%%  \caption{MRFy run-time performance with varying
%%  numbers of processing cores on a large protein}
%%  \label{multicore}
%%  \end{figure}

\begin{figure}[h!]
\centerline{\includegraphics{speedup}}
\bigskip
\caption{\mrfy's parallel speedup}
\label{fig:speedup}
\end{figure}

\begin{figure}[h!]
\centerline{\includegraphics{efficiency}}
\bigskip
\caption{\mrfy's parallel efficiency}
\label{fig:efficiency}
\end{figure}
